Liveness-Based Garbage Collection for Lazy Languages

### 2015-06-04 paper/paper.tex
liveness

\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it
      %probably doesn't scale anyway.
\newcommand{\mycomment}[1]{}
\definecolor{Myblue}{rgb}{.2,0,1}
\newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
\newcommand{\cred}[1]{{\color{red}{(#1)}}}
\newcommand{\alan}[1]{{\color{red}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\blankout}[1]{}

@@

The  solution  is  to   also  treat  the  operations  mentioned  above
symbolically.  We  introduce three  new  symbols.   First, \bcar\  and
\bcdr\ are defined as:
\begin{align*}
 \bcar\sigma &\triangleq \{\alpha \mid 0\alpha \in \sigma\}\\
 \bcdr\sigma &\triangleq \{\alpha \mid 1\alpha \in \sigma\}
\end{align*}
and \cbar\ as:
\begin{align*}
 \cbar\sigma \triangleq& \emptyset~\mbox{if}~\sigma = \emptyset\\
                       & \{\epsilon\} ~\mbox{otherwise}
\end{align*}

We can now rewrite the \CONS\ and the \CAR\ rules of $\mathit{ref}$ as:
\begin{align*}
&\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
= x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons},~
\mbox{and} \\
&\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          =   x.\cbar\sigma \cup x.0\sigma
\end{align*}
and the \Lfunonly\ rule
for \SIF\ as:
\begin{align*}
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) =
                    &\mathcal{L}(e_1,\sigma,\Lfonly)~\uplus
        \mathcal{L}(e_2,\sigma,\Lfonly)~\uplus\\
        &[\,\mathit{ep} \mapsto  x.\cbar \sigma]
\end{align*}
The rules for \CDR, \PRIM\ and  \NULLQ\  are also modified in a manner
similar to \CAR.
The  defintions of \bcar\, \bcdr\ and \cbar\  induce   the  following
relation $\hookrightarrow$ over sets of access paths:
\begin{align*}
  &\sigma_1\bcar\sigma_2  \hookrightarrow
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \acar\sigma_2',\\
&\sigma_1\bcdr\sigma_2  \hookrightarrow
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
  \acdr\sigma_2',~\mbox{and}\\
&\sigma_1\cbar \emptyset \hookrightarrow \emptyset
\end{align*}

@@

\begin{figure*}[t!]
  \begin{tabular}{cc}
    \begin{minipage}{.45\textwidth}
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\
(\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\
$\psi_1\!\!:\,$ \px
            \UNL{4} \hspace*{.27cm} $\pi_3\!\!:\,
            $(\LET\ \pv\ $\leftarrow 0$ \IN
            \UNL{5} \hspace*{.32cm} $\pi_4\!\!:\,
(\SRETURN~\psi_2:\pv)$
            \UNL{4} \hspace*{.29cm}    $\pi_5\!\!:\, $(\LET~\pu\
$\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{.34cm}   $\pi_6\!\!:\, $(\LET~\py\
$\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{.34cm} $\pi_7\!\!:\,
            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
            \UNL{7} \hspace*{.34cm} $\pi_8\!\!:\,
(\SRETURN~\psi_3:\pz)$)))))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $(\DEFINE\ (\main)$
          \UNL{2} \!\!$\pi_9\!\!:\, (\LET\  \pa\  \leftarrow$ \ldots
\IN
	  \UNL{3}   \hspace*{.05cm}$\pi_{10}\!\!:\,      $ (\LET\ \pb\
$\leftarrow$  \ldots \IN
          \UNL{4}   \hspace*{.15cm}    $\pi_{11}\!\!:\,
          $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
          \UNL{5}  \hspace*{.25cm}  $\pi_{12}\!\!:\,
(\SRETURN~\psi_4:\pw)))))$
\end{uprogram}}
        %%}
    \end{minipage}

@@

        \small
\begin{eqnarray*}
        \mathcal{L}(e_{\length},\sigma,\Lfonly) &=&
              [\;ep_1 \mapsto \lbrace \px.\sigma_{\length},
\xl.\cbar\sigma_{\length} \rbrace ,~ep_2 \mapsto \lbrace \rbrace,\\
 &&          \;\;  ep_3 \mapsto \lbrace \pz.\sigma_{\length},
                \py.\cbar\sigma_{\length},
\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}},\\
&&            \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;
\xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}
                \rbrace~\rbrack \\
            \cal{P_{\mbox{\tiny \length}}} &=&
                [\;\pi_1 \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace,  \
                 \pi_2 \mapsto \lbrace ep_1, ep_2, ep_3
                  \rbrace, \\
&&\;\;  \pi_3 \mapsto \lbrace ep_2
                  \rbrace ,  \pi_4 \mapsto \lbrace ep_2
                  \rbrace, \pi_5 \mapsto \lbrace  ep_3 \rbrace, \\
 &&  \;\;                        \pi_6 \mapsto \lbrace ep_3 \rbrace,
\pi_7 \mapsto
                  \lbrace ep_3\rbrace, \pi_8 \mapsto \lbrace ep_3
\rbrace~\rbrack\\
              \mathcal{L}(e_{\main},\sigma,\Lfonly) &=&
                [\;
                  ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}},
\pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots\rbrace
                  ]\\
              \cal{P_{\mbox{\tiny \main}}} &=&
                  \lbrack~\pi_6 \mapsto \lbrace ep_4 \rbrace,  \
                   \pi_7 \mapsto \lbrace ep_4 \rbrace,  \
                  \pi_8 \mapsto \lbrace ep_4 \rbrace~\rbrack
\end{eqnarray*}
    \end{minipage} \\
(a)&
(b)
    %% \end{ tabular}}
  \end{tabular}
%\end{picture}
\kern -3ex
%\vspace{0.5in}
\caption{(a) Example program and (b) its liveness
maps.}\label{fig:mot-example2}
\end{figure*}
%================================================================

@@

\subsection{Generating Liveness Equations}
The  program  in  Figure~\ref{fig:mot-example2}  serves as  a  running
example.   Given a  function $\mathit{f}$,  we now  describe  how to
generate    equations    for     the    demand    transformation
\Lfonly$_\mathit{f}$.   Starting with a  symbolic demand  $\sigma$, we
determine  \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.  In  particular, we
consider $\Lanv{1}{x_i}$,  the  liveness of    the
$i^{\mbox{\footnotesize  th}}$
parameter $x_i$ at the  program point at the  beginning of
$e_{\mathit{f}}$ (assumed  to be  $\pi_1$). By the rule {\sc
  live-define}, this should be the
same as $\Lf{f}{i}{\sigma}$. Applying this to  \length, we have:
\begin{eqnarray*}
&& \Lanv{1}{\xl} = \cbar\sigma \cup \acdr\Lf{\length}{1}{\cbar\sigma}
\end{eqnarray*}
and  the only equation defining \Lfone{\length} is:
  \begin{eqnarray*}
   && \Lf{\length}{1}{\sigma}
    =  \cbar\sigma \cup \acdr\Lf{\length}{1}{\cbar\sigma}
   \end{eqnarray*}

In general, the equations for  \Lfonly\ are recursive since, as in the
case  of  \length,  \Lan{1}\  may   have  been  defined  in  terms  of
\Lfone{\mathit{f}}.  However,  it is desirable  to have a  closed form
solution for $\Lfone{\mathit{f}}$.  As mentioned in \cite{asati14lgc},
each  of the liveness  rules modifies  a demand  by prefixing  it with
symbols  in the  alphabet  $\lbrace \acar,  \acdr,\bcar, \bcdr,  \cbar
\rbrace$, and therefore we can assume that $\Lf{f}{i}{\sigma}$ has the
closed form:
\begin{eqnarray}
\label{eq:LF:DI}
  \Lf{f}{i}{\sigma} = \Df{f}{i}.\sigma
\end{eqnarray}
where \Df{f}{i} are sets of  strings over the alphabet mentioned above.
  Substituting the guessed form in
the  equation  describing  \Lfonly$_{\mathit  f}$, and  factoring  out
$\sigma$,  we get  an equation  for \Df{f}{i}  that is  independent of
$\sigma$.   Any   solution  for   \Df{f}{i}  yields  a   solution  for
\Lfonly$_{\mathit  f}$.  Applied  to  \Lfonly$_{\mathit \length}$,  we
get:
  \begin{eqnarray*}
  &&  \Df{\length}{1}.\sigma = \cbar.\sigma
    \cup \acdr\Df{\length}{1}(\cbar\sigma), ~\mbox{or}\\
  &&  \Df{\length}{1} = \lbrace\cbar\rbrace
    \cup \acdr\Df{\length}{1}\cbar
  \end{eqnarray*}

Note that this equation can also be viewed as a CFG with \{\acdr,
\cbar\} as the terminal symbols and \Df{\length}{1} as the sole
non-terminal.

@@

To avoid analysing the body of a function for each call, we calculated
the liveness  for the arguments and  the variables in  a function with
respect to a symbolic demand  $\sigma$.  To get the actual liveness we
calculate an over-approximation of the  actual demands made by all the
calls by a {\em summary demand} and calculate the liveness at each GC
point inside the  function based on this summary  demand.  Even though
we  lose precision  due to  this, our  experiments show  that  this is
acceptable  \comment{Is  this  claim  true?}.   The  0-CFA-style  {\em
  summary demand}  is calculated by taking  a union of  the demands at
every call site of a function.

Consider a function $g$ containing a  call to $f$ at a site $\pi$, say
$\pi\!\!:\!(\LET~x   \leftarrow   (f\,y_1\,\ldots\,y_n)~\IN~\pi_i:e)$.
Let the  demand on $g$  be $\sigma_g$ and,  based on this  demand, the
liveness of  $x$ at $\pi_i$ be  $\Lanv{i}{x}$.  By the  $\LET$ rule of
Figure~\ref{fig:live-judge},    the   call   at    $\pi$   contributes
$\Lanv{i}{x}$ to  the demand $\sigma_f$.  Assuming that  there are $k$
call sites to function $f$, $\pi^1$ (in function $g^1$) \ldots $\pi^k$
(in   function  $g^k$),  the   over-approximation  of   $\sigma_f$  is
$\deltacall{f}{\pi^1}{g^1} \cup \cdots \cup \deltacall{f}{\pi^k}{g^k}$.
The distinguished function  \mainpgm\ is a special case.  We assume it
is called through a  printing mechanism with demand $\sigma_\mainpgm =
\{\acar,\acdr\}^\ast$ (denoted $\sigma_{\!all}$).

For  the  running  example,   $\length$  has  calls  from  $\main$  at
$\pi_{10}$ and a recursive call at $\pi_6$.
%\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
So $\sigma_{\length} =
     \deltacall{\length}{\pi_{11}}{\main}  \cup
\deltacall{\length}{\pi_6}{\length}$.
Filling in  the values gives:
\begin{eqnarray*}
\sigma_{\length}    &=&
 \sigma_{\!all}  ~\cup~
\acdr\Lf{\length}{1}{\cbar\sigma_{\mathit{\length}}}
\end{eqnarray*}
For some of the program points in \length\  and \main,  this gives
{\color {red} Prasanna, please fill this correctly}. You could use the
set notation if the map notation spills over:
\begin{align*}
\Lan{1} &= \lbrace  \px.\sigma_{\length}, \xl.\cbar\sigma_{\length},
\pz.\sigma_{\length},\
              \py.\cbar\sigma_{\length}, \\
&~~~~~~~\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}},
\xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace \\
\Lan{2} & =  \lbrace  \px.\sigma_{\length}, \xl.\cbar\sigma_{\length},
\pz.\sigma_{\length},\
                  \py.\cbar\sigma_{\length}, \\
&\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}},
\xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace \\
\Lan{5} &= \lbrace \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}},
\xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}\rbrace \\
\Lan{8} &= \lbrace \pw.\sigma_{\mathit{\!all}},
\pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace
\end{align*}

In summary, the equations generated during liveness analysis are:
\begin{enumerate}
\item   For    each   function   $\mathit{f}$,    equations   defining
\Df{f}{i} for use by \Lfonly$_{\mathit f}$.
\item  For  each   function  $\mathit{f}$, an  equation  defining
the summary demand  $\sigma_{\mathit f}$ on $e_f$.
% \marginpar{Consistency: summary vs cumulative?}
\item For each function $\mathit{f}$ (including $\mainpgm$ for
  $e_\mainpgm$) \comment{Have we defined GC points for lazy languages?
No, we define it only in the implementation section as of now.}
an equation defining
  liveness at each GC point of $e_{\mathit f}$.

\subsection{Solving liveness equations---the grammar
interpretation}\label{sec:grammar-formulation}
The  equations above can now be re-interpreted as a
context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
\bcar, \bcdr, \cbar\rbrace$.  Let  \var{$X$} denote the non-terminal
for
a variable  $X$ occurring on  the LHS of the  equations generated
from the analysis.  We can  think of the resulting productions as
being associated with several  grammars, one for each non-terminal
\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
grammar    for   \var{\Lanv{5}{\xl}}   comprises    the   following
productions:
 \begin{eqnarray*}
\var{\Lanv{5}{\xl}}  &\rightarrow&   \var{\acdr \Lf
  {\length}{1}{\cbar\sigma_{\length}}} \mid \cbar\sigma_{\length} \\
\end{eqnarray*}
\comment{Pick up a richer equation or more than one liveness equations}
Other equations can be converted  similarly.     The  language
generated  by  \var{\Lanv{i}{x}},
denoted $\mathscr{L}(\var{\Lanv{i}{x}})
$,  is the desired  solution of
\Lanv{i}{x}.
However, recall from  our earlier  discussion that
the  decision problem that  we are  interested in  during garbage
collection is:
\begin{quote}
Let $x.\alpha$ be  a forward  access path  consisting  only  of   edges
\acar\  and \acdr\  (but  not \bcar\  or  \bcdr).   Let
$\mathscr{L}(\var{\Lanv{i}{x}})
\stackrel{*}{\hookrightarrow} \sigma$, where  $\sigma$  consists
of forward paths only. Then does $\alpha \in \sigma$?
\end{quote}
We could convert  the  rules  defining  $\hookrightarrow$  into
productions and  add them to the grammar. However, this results
in  an {\em  unrestricted grammar}~\cite{hopcraft90toc},  and the
membership  problem   for  such  grammars   is
undecidable. \comment{Rewrite it to say that this is a CFG with 3
  unrestricted rules and the membership for even this grammar is
undecidable}
\comment{State the undecidability of the decision problem above.}

@@

Generated by bibtex from your ~.bib file.  Run latex,
then bibtex, then latex twice (to resolve references)
to create the ~.bbl file.  Insert that ~.bbl file into
the .tex source file and comment out
the command \texttt{{\char'134}thebibliography}.
% This next section command marks the start of
% Appendix B, and does not continue the present hierarchy
\section{More Help for the Hardy}
The sig-alternate.cls file itself is chock-full of succinct
and helpful comments.  If you consider yourself a moderately
experienced to expert user of \LaTeX, you may find reading
it useful but please remember not to change it.
%\balancecolumns % GM June 2007
% That's all folks!
\end{enumerate}

>>>>

\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it
      %probably doesn't scale anyway.
\newcommand{\mycomment}[1]{}
\definecolor{Myblue}{rgb}{.2,0,1}
\newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
\newcommand{\cred}[1]{{\color{red}{#1}}}
\newcommand{\alan}[1]{{\color{red}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\blankout}[1]{}

@@

The  solution  is  to   also  treat  the  operations  mentioned  above
symbolically.  We  introduce three  new  symbols.   First, \bcar\  and
\bcdr\ are defined as:
\begin{align*}
 \bcar\sigma &\triangleq \{\alpha \mid 0\alpha \in \sigma\}\\
 \bcdr\sigma &\triangleq \{\alpha \mid 1\alpha \in \sigma\}
\end{align*}
and \cbar\ as:
\begin{align*}
 \cbar\sigma \triangleq\; & \emptyset~\mbox{if}~\sigma = \emptyset\\
                       & \{\epsilon\} ~\mbox{otherwise}
\end{align*}
Note that \cbar\ is idempotent. 
We can now rewrite the \CONS\ and the \CAR\ rules of $\mathit{ref}$ as:
\begin{align*}
&\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
= x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons},~
\mbox{and} \\
&\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          =   x.\cbar\sigma \cup x.0\sigma
\end{align*}
and the \Lfunonly\ rule
for \SIF\ as:
\begin{align*}
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) =
                    &\mathcal{L}(e_1,\sigma,\Lfonly)~\uplus
        \mathcal{L}(e_2,\sigma,\Lfonly)~\uplus\\
        &[\,\mathit{ep} \mapsto  x.\cbar \sigma]
\end{align*}
The rules for \CDR, \PRIM\ and  \NULLQ\  are also modified in a manner
similar to \CAR.
The  defintions of \bcar\, \bcdr\ and \cbar\  induce   the  following
relation $\hookrightarrow$ over sets of access paths:
\begin{align*}
  &\sigma_1\bcar\sigma_2  \hookrightarrow
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \acar\sigma_2',\\
&\sigma_1\bcdr\sigma_2  \hookrightarrow
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
  \acdr\sigma_2',~\mbox{and}\\
&\sigma_1\cbar \emptyset \hookrightarrow \emptyset
\end{align*}

@@

\begin{figure*}[t!]
  \begin{tabular}{cc}
    \begin{minipage}{.40\textwidth}
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\
(\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\
$\psi_1\!\!:\,$ \px
            \UNL{4} \hspace*{.27cm} $\pi_3\!\!:\,
            $(\LET\ \pv\ $\leftarrow 0$ \IN
            \UNL{5} \hspace*{.32cm} $\pi_4\!\!:\,
(\SRETURN~\psi_2:\pv)$
            \UNL{4} \hspace*{.29cm}    $\pi_5\!\!:\, $(\LET~\pu\
$\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{.34cm}   $\pi_6\!\!:\, $(\LET~\py\
$\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{.34cm} $\pi_7\!\!:\,
            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
            \UNL{7} \hspace*{.34cm} $\pi_8\!\!:\,
(\SRETURN~\psi_3:\pz)$)))))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $(\DEFINE\ (\main)$
          \UNL{2} \!\!$\pi_9\!\!:\, (\LET\  \pa\  \leftarrow$ \ldots
\IN
	  \UNL{3}   \hspace*{.05cm}$\pi_{10}\!\!:\,      $ (\LET\ \pb\
$\leftarrow$  \ldots \IN
          \UNL{4}   \hspace*{.15cm}    $\pi_{11}\!\!:\,
          $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
          \UNL{5}  \hspace*{.25cm}  $\pi_{12}\!\!:\,
(\SRETURN~\psi_4:\pw)))))$
\end{uprogram}}
        %%}
    \end{minipage}

@@

        \small
\begin{eqnarray*}
        \mathcal{L}(e_{\length},\sigma,\Lfonly) &=&
              [\;ep_1 \mapsto \lbrace \px.\cbar \sigma,
\xl.\cbar\sigma\rbrace ,~ep_2 \mapsto \lbrace \cred{
\pv.\sigma}  \rbrace,\\
 &&          \;\;  ep_3 \mapsto \lbrace \pz.\sigma,
                \py.\cbar\sigma,
\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma},\\
&&            \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\; 
\xl.\{ \acdr\Lf{\length}{\mbox{1}}{\cbar\sigma} \cup 
%%\;\;\;\;\;\;\; \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\\
\cred{\cbar\Lf{\length}{1}{\cbar\sigma}}\}\}
             \rbrack \\
            \cal{P_{\mbox{\tiny \length}}} &=&
                [\;\pi_1 \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace,  \
                 \pi_2 \mapsto \lbrace ep_1, ep_2, ep_3
                  \rbrace, \\
&&\;\;  \pi_3 \mapsto \lbrace ep_2
                  \rbrace ,  \pi_4 \mapsto \lbrace ep_2
                  \rbrace, \pi_5 \mapsto \lbrace  ep_3 \rbrace, \\
 &&  \;\;                        \pi_6 \mapsto \lbrace ep_3 \rbrace,
\pi_7 \mapsto
                   \lbrace ep_3\rbrace, \pi_8 \mapsto \lbrace ep_3
\rbrace~\rbrack\\
              \mathcal{L}(e_{\main},\sigma,\Lfonly) &=&
                [\;
                  ep_4 \mapsto \lbrace \pw.\sigma,
           \pb.\Lf{\length}{\mbox{1}}{\sigma}, \ldots\rbrace
                   ] \\
               \cal{P_{\mbox{\tiny \main}}} &=&
                   \lbrack~\pi_9 \mapsto \lbrace ep_4 \rbrace,  \
                    \pi_{10} \mapsto \lbrace ep_4 \rbrace,  \
                   \pi_{11} \mapsto \lbrace ep_4 \rbrace,  \pi_{12} \mapsto \lbrace ep_4 \rbrace~\rbrack
\end{eqnarray*}
    \end{minipage} \\
(a)&
(b)
    %% \end{ tabular}}
  \end{tabular}
%\end{picture}
\kern -3ex
%\vspace{0.5in}
\caption{(a) Example program and (b) its liveness
maps.}\label{fig:mot-example2}
\end{figure*}
%================================================================

@@

\subsection{Generating Liveness Equations}
The  program  in  Figure~\ref{fig:mot-example2}  serves as  a  running
example.   Given a  function $\mathit{f}$,  we now  describe  how to
generate    equations    for     the    demand    transformation
\Lfonly$_\mathit{f}$.   Starting with a  symbolic demand  $\sigma$, we
determine  \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.  In  particular, we
consider $\Lanv{1}{x_i}$,  the  liveness of    the
$i^{\mbox{\footnotesize  th}}$
parameter $x_i$ at the  program point at the  beginning of
$e_{\mathit{f}}$ (assumed  to be  $\pi_1$). By the rule {\sc
  live-define}, this should be the
same as $\Lf{f}{i}{\sigma}$. Applying this to  \length, we have:
\begin{eqnarray*}
&& \Lanv{1}{\xl} = \cbar\sigma \cup \acdr\Lf{\length}{1}{\cbar\sigma}
  \cup \cbar\Lf{\length}{1}{\cbar\sigma}
\end{eqnarray*}
and  the only equation defining \Lfone{\length} is:
  \begin{eqnarray*}
   && \Lf{\length}{1}{\sigma}
    =  \cbar\sigma \cup \acdr\Lf{\length}{1}{\cbar\sigma} \cup \cbar\Lf{\length}{1}{\cbar\sigma}
\end{eqnarray*}

In general, the equations for  \Lfonly\ are recursive since, as in the
case  of  \length,  \Lan{1}\  may   have  been  defined  in  terms  of
\Lfone{\mathit{f}}.  However,  it is desirable  to have a  closed form
solution for $\Lfone{\mathit{f}}$.  As mentioned in \cite{asati14lgc},
each  of the liveness  rules modifies  a demand  by prefixing  it with
symbols  in the  alphabet  $\lbrace \acar,  \acdr,\bcar, \bcdr,  \cbar
\rbrace$, and therefore we can assume that $\Lf{f}{i}{\sigma}$ has the
closed form:
\begin{eqnarray}
\label{eq:LF:DI}
  \Lf{f}{i}{\sigma} = \Df{f}{i}.\sigma
\end{eqnarray}
where \Df{f}{i} are sets of strings over the alphabet mentioned above.
Substituting   the   guessed   form   in   the   equation   describing
\Lfonly$_{\mathit f}$, and factoring  out $\sigma$, we get an equation
for  \Df{f}{i} that  is  independent of  $\sigma$.   Any solution  for
\Df{f}{i}  yields a  solution for  \Lfonly$_{\mathit f}$.   Applied to
\Lfonly$_{\mathit \length}$, we get:
  \begin{eqnarray*}
&&  \Lf{\length }{1}{\sigma} = \Df{\length}{1}.\sigma,~\mbox{and}\\
&&   \Df{\length}{1} = \cbar \cup \acdr\Df{\length}{1}\cbar 
       \cup \cbar\Df{\length}{1}.{\cbar}
  \end{eqnarray*}

Note that this equation can also be viewed as a CFG with \{\acdr,
\cbar\} as terminal symbols and \Df{\length}{1} as the sole
non-terminal.

@@

To avoid analyzing the body of a function for each call, we calculated
the liveness  for the arguments and  the variables in  a function with
respect to a symbolic demand  $\sigma$.  To get the actual liveness we
calculate an over-approximation of the  actual demands made by all the
calls by a {\em summary demand} and calculate the liveness at each GC
point \comment{have we defined what a garbage collection point
  is?} inside the  function based on this summary  demand.  Even though
we  lose precision  due to  this, our  experiments show  that  this is
acceptable  \comment{Is  this  claim  true?}.   The  0-CFA-style  {\em
  summary demand}  is calculated by taking  a union of  the demands at
every call site of a function.

Consider a function $g$ containing a  call to $f$ at a site $\pi$, say
$\pi\!\!:\!(\LET~x   \leftarrow   (f\,y_1\,\ldots\,y_n)~\IN~\pi_i:e)$.
Let the  demand on $g$  be $\sigma_g$ and,  based on this  demand, the
liveness of  $x$ at $\pi_i$ be  $\Lanv{i}{x}$.  By the  $\LET$ rule of
Figure~\ref{fig:live-judge},    the   call   at    $\pi$   contributes
$\Lanv{i}{x}$ to  the demand $\sigma_f$. Let us denote the
contribution of a call site $\pi$ in a function $g$ to the overall
demand on the function $f$ as $\deltacall{f}{\pi}{g}$. Assuming that  there are $k$
call sites to function $f$, $\pi^1$ (in function $g^1$) \ldots $\pi^k$
(in   function  $g^k$),  the   over-approximation  of   $\sigma_f$  is
$\deltacall{f}{\pi^1}{g^1} \cup \cdots \cup \deltacall{f}{\pi^k}{g^k}$.
The distinguished function  \mainpgm\ is a special case.  We assume it
is called through a  printing mechanism with demand $\sigma_\mainpgm =
\{\acar,\acdr\}^\ast$ (denoted $\sigma_{\!all}$).

For  the  running  example,   $\length$  has  calls  from  $\main$  at
$\pi_{10}$ and a recursive call at $\pi_6$.
So $\sigma_{\length} =
     \deltacall{\length}{\pi_{11}}{\main}  \cup 
\deltacall{\length}{\pi_6}{\length}$.
Filling in  the values gives:
\begin{eqnarray*}
\sigma_{\length}    &=&
 \sigma_{\!all}  ~\cup~{\cbar\sigma_{\mathit{\length}}}
\end{eqnarray*}
As examples, the liveness of \Lanv{1}{\pl} and \Lanv{9}{\pb} in terms
of  $\sigma_{\length}$ are:
\begin{align*}
\Lanv{1}{\pl} &= \cbar\sigma_{\length} \cup \acdr\Lf{\length}{1}{\cbar\sigma_{\length}}
  \cup \cbar\Lf{\length}{1}{\cbar\sigma_{\length}}
 \\
\Lanv{9}{\pb} &= \Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}
\end{align*}

In summary, the equations generated during liveness analysis are:
\begin{enumerate}
\item   For    each   function   $\mathit{f}$,    equations   defining
\Df{f}{i} for use by \Lfonly$_{\mathit f}$.
\item  For  each   function  $\mathit{f}$, an  equation  defining
the summary demand  $\sigma_{\mathit f}$ on $e_f$.
\item For each function $\mathit{f}$ (including $\mainpgm$ for
  $e_\mainpgm$) 
an equation defining
  liveness at each GC point of $e_{\mathit f}$.
\end{enumerate}
\subsection{Solving liveness equations---the grammar
interpretation}\label{sec:grammar-formulation}
The  equations above can now be re-interpreted as a
context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
\bcar, \bcdr, \cbar\rbrace$.  Let  \var{$X$} denote the non-terminal
for
a variable  $X$ occurring on  the LHS of the  equations generated
from the analysis.  We can  think of the resulting productions as
being associated with several  grammars, one for each non-terminal
\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
grammar    for   \var{\Lanv{5}{\xl}}   comprises    the   following
productions:
 \begin{eqnarray*}
\var{\Lanv{5}{\xl}}  &\rightarrow&   \var{\acdr \Lf
  {\length}{1}{\cbar\sigma_{\length}}} \mid \cbar\sigma_{\length} \\
\end{eqnarray*}
\comment{Pick up a richer equation or more than one liveness equations}
Other equations can be converted  similarly.     The  language
generated  by  \var{\Lanv{i}{x}},
denoted $\mathscr{L}(\var{\Lanv{i}{x}})
$,  is the desired  solution of
\Lanv{i}{x}.
However, recall from  our earlier  discussion that
the  decision problem that  we are  interested in  during garbage
collection is:
\begin{quote}
Let $x.\alpha$ be  a forward  access path  consisting  only  of   edges
\acar\  and \acdr\  (but  not \bcar\  or  \bcdr).   Let
$\mathscr{L}(\var{\Lanv{i}{x}})
\stackrel{*}{\hookrightarrow} \sigma$, where  $\sigma$  consists
of forward paths only. Then does $\alpha \in \sigma$?
\end{quote}
We could convert  the  rules  defining  $\hookrightarrow$  into
productions and  add them to the grammar. However, this results
in  an {\em  unrestricted grammar}~\cite{hopcraft90toc},  and the
membership  problem   for  such  grammars   is
undecidable. \comment{Rewrite it to say that this is a CFG with 3
  unrestricted rules and the membership for even this grammar is
undecidable}
\comment{State the undecidability of the decision problem above.}

@@

Generated by bibtex from your ~.bib file.  Run latex,
then bibtex, then latex twice (to resolve references)
to create the ~.bbl file.  Insert that ~.bbl file into
the .tex source file and comment out
the command \texttt{{\char'134}thebibliography}.
% This next section command marks the start of
% Appendix B, and does not continue the present hierarchy
\section{More Help for the Hardy}
The sig-alternate.cls file itself is chock-full of succinct
and helpful comments.  If you consider yourself a moderately
experienced to expert user of \LaTeX, you may find reading
it useful but please remember not to change it.
%\balancecolumns % GM June 2007
% That's all folks!
